---
title: "Ecotourism Tutorial"
subtitle: "Joining occurrence, weather, and tourism data"
author: "Mahjabin Siddika Oyshi"
format:
  html:
    embed-resources: true
---
```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false

library(ecotourism)
library(tidyverse)

data(orchids)
data(weather)
data(tourism_quarterly)
data(tourism_region)
tourism_total <- tourism_quarterly |>
  dplyr::group_by(ws_id, year, quarter) |>
  dplyr::summarise(
    total_trips = sum(trips, na.rm = TRUE),
    .groups = "drop"
  )
```
## Objectives

In this tutorial you will work with orchid sightings from the
`ecotourism` package, and link them to weather and tourism data.
We will focus on Western Australia during September, when many orchids
are flowering.

By the end you should be able to:

- filter occurrence records by state and month  
- join occurrence data to daily weather using `ws_id` and `date`  
- join occurrence data to quarterly tourism counts  
- check how complete the joined data is  

## Preparation

We will use four linked datasets:

- `orchids`: orchid occurrence records (2014–2024)  
- `weather`: daily weather by station  
- `tourism_quarterly`: quarterly domestic visitor counts by region  
- `tourism_region`: tourism region names with coordinates and `ws_id`  

## Exercises

### Question 1 — Western Australia, September

In this question you will look at orchid sightings in Western Australia
during September.

1. Filter `orchids` to rows where `obs_state == "Western Australia"`
   and `month == 9`.  
2. Join these rows to `weather` using `ws_id` and `date`.  
3. Count  
   - how many sightings you have in total  
   - how many of them have a non‑missing `max`
     (daily maximum temperature)  

What does this tell you about how complete the weather data is for
these sightings?

#### Solution 1

```{r}
#| label: q1-solution
#| echo: TRUE

wa_sep <- orchids |>
  dplyr::filter(
    obs_state == "Western Australia",
    month == 9
  )

wa_sep_weather <- wa_sep |>
  dplyr::left_join(weather, by = c("ws_id", "date"))

# total September sightings in WA
nrow(wa_sep)

# sightings with weather data
sum(!is.na(wa_sep_weather$max))

# proportion missing max temperature
mean(is.na(wa_sep_weather$max))
```
Only about 7% of September orchid sightings in Western Australia have a
recorded maximum temperature, so the weather data is very incomplete.
This suggests gaps in the weather records rather than a lack of orchid
activity, and shows why updating the weather part of the package is
important.

### Question 2 — Adding tourism

Now bring in the tourism data.

1. Start from the Western Australia September orchid data you used in
   Question 1.
2. Join it to the `tourism_total` table using `ws_id`, `year`, and
   `quarter`. Remember that September is in quarter 3.
3. For Western Australia, report:
   - the total number of September orchid sightings
   - the average `total_trips`
   - the proportion of sightings that have a non‑missing `max`
     temperature.

What do these numbers suggest about how much tourism and weather
information is available for the places where orchids are being
recorded?

#### Solution 2

```{r}
#| label: q2-solution
#| echo: TRUE

# Western Australia, September orchids with quarter
wa_sep_q <- orchids |>
  dplyr::filter(
    obs_state == "Western Australia",
    month == 9
  ) |>
  dplyr::mutate(quarter = 3L)

# join tourism totals
wa_sep_tour <- wa_sep_q |>
  dplyr::left_join(
    tourism_total,
    by = c("ws_id", "year", "quarter")
  )

# join weather again so we still have max temperature
wa_sep_full <- wa_sep_tour |>
  dplyr::left_join(weather, by = c("ws_id", "date"))

summary_q2 <- wa_sep_full |>
  dplyr::summarise(
    n_sightings   = dplyr::n(),
    mean_trips    = mean(total_trips, na.rm = TRUE),
    prop_with_max = mean(!is.na(max))
  )

summary_q2
```
There are 14,911 September orchid sightings in Western Australia, but
the average tourism trip count at those stations is only about 33. Just
under 7% of sightings have a recorded maximum temperature. This tells
us that most orchid records sit in areas with relatively low reported
tourism, and weather data is still very patchy for these locations.

### Question 3 — Sightings vs tourism

Using the joined data from Question 2:

1. For each year in Western Australia, work out:
   - total September orchid sightings
   - average `total_trips`.
2. Make a scatter plot with `total_trips` on the x‑axis and total
   orchid sightings on the y‑axis.

Does it look like years with more visitors also have more recorded
orchid sightings? How might observer effort explain any pattern you
see?

#### Solution 3

```{r}
#| label: q3-solution
#| echo: TRUE
#| fig-width: 6
#| fig-height: 4

# Western Australia, September orchids with quarter
wa_sep_q <- orchids |>
  dplyr::filter(
    obs_state == "Western Australia",
    month == 9
  ) |>
  dplyr::mutate(quarter = 3L)

# join tourism totals
wa_sep_tour <- wa_sep_q |>
  dplyr::left_join(
    tourism_total,
    by = c("ws_id", "year", "quarter")
  )

# (optional) join weather to keep using all three tables
wa_sep_full <- wa_sep_tour |>
  dplyr::left_join(weather, by = c("ws_id", "date"))

# yearly totals for WA, September
yearly_wa <- wa_sep_tour |>
  dplyr::group_by(year) |>
  dplyr::summarise(
    n_sightings = dplyr::n(),
    mean_trips  = mean(total_trips, na.rm = TRUE),
    .groups = "drop"
  )

yearly_wa
ggplot(yearly_wa,
       aes(x = mean_trips, y = n_sightings)) +
  geom_point() +
  labs(
    x = "Average quarterly tourism trips (WA, September)",
    y = "Total orchid sightings (WA, September)",
    title = "September orchid sightings vs tourism in Western Australia"
  ) +
  theme_minimal()
```
Overall, there is a loose positive relationship between tourism and orchid sightings: years with higher average trip counts often have more recorded September sightings in Western Australia.