---
title: "Ecotourism Tutorial"
subtitle: "Joining occurrence, weather, and tourism data"
author: "Mahjabin Siddika Oyshi"

format:
  html:
    embed-resources: true
    code-fold: true
    code-tools: true
    toc: true
    theme: cosmo
---

```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false
library(ecotourism)
library(tidyverse)
library(plotly)
library(DT)
library(leaflet)

data(orchids)
data(weather)
data(tourism_quarterly)
data(tourism_region)

tourism_total <- tourism_quarterly |>
  group_by(ws_id, year, quarter) |>
  summarise(
    total_trips = sum(trips, na.rm = TRUE),
    .groups = "drop"
  )
```

## Objectives

In this tutorial you will work with orchid sightings from the `ecotourism` package, and link them to weather and tourism data. We will focus on Western Australia during September, when many orchids are flowering.

By the end you should be able to:

-   filter occurrence records by state and month\
-   join occurrence data to daily weather using `ws_id` and `date`\
-   join occurrence data to quarterly tourism counts\
-   check how complete the joined data is

## Preparation

**Setup note**\
If you don’t have the package installed yet, run in a fresh R session:

``` r
# install.packages("pak")
pak::pak("vahdatjavad/ecotourism")
```

We will use four linked datasets:

-   `orchids`: orchid occurrence records (2014–2024)\
-   `weather`: daily weather by station\
-   `tourism_quarterly`: quarterly domestic visitor counts by region\
-   
    -   `tourism_quarterly`: quarterly domestic visitor counts by region\\
-   `tourism_region`: tourism region names with coordinates and `ws_id`

## Explore the orchid dataset

```{r}
#| echo: false
datatable(
  orchids,
  options = list(pageLength = 8)
)
```

Map of Orchid Sightings

```{r}
#| echo: false
orchids_map <- orchids |>
  filter(!is.na(obs_lon), !is.na(obs_lat)) |>
  sample_n(500)

leaflet(orchids_map) |>
  addTiles() |>
  addCircleMarkers(
    lng = ~obs_lon,
    lat = ~obs_lat,
    radius = 3,
    popup = ~paste0(
      "Species: ", sci_name,
      "<br>Date: ", date,
      "<br>State: ", obs_state
    )
  )

```

## Exercises

### Question 1 — Western Australia in September

In this question you will look at orchid sightings in Western Australia during September.

1.  Filter `orchids` to rows where `obs_state == "Western Australia"` and `month == 9`.\
2.  Join these rows to `weather` using `ws_id` and `date`.\
3.  Count
    -   how many sightings you have in total\
    -   how many of them have a non‑missing `max` (daily maximum temperature)

What does this tell you about how complete the weather data is for these sightings?

#### Solution 1

```{r}
#| label: q1-solution
#| code-fold: true

wa_sep <- orchids |>
  filter(obs_state == "Western Australia",
         month == 9)

wa_sep_weather <- wa_sep |>
  left_join(weather, by = c("ws_id", "date"))

n_total <- nrow(wa_sep)
n_weather <- sum(!is.na(wa_sep_weather$max))

n_total
n_weather
mean(is.na(wa_sep_weather$max))
```

Data completeness visualization

```{r}
#| echo: false

wa_sep_weather |>
  mutate(weather_available = !is.na(max)) |>
  count(weather_available) |>
  plot_ly(
    labels = ~weather_available,
    values = ~n,
    type = "pie"
  )
```

Only about 7% of September orchid sightings in Western Australia have a recorded maximum temperature (the pie chart shows that roughly 93% are missing), so the weather data is very incomplete. This suggests gaps in the weather records rather than a lack of orchid activity, and shows why updating the weather part of the package is important.

### Question 2 — Adding tourism

Now bring in the tourism data.

1.  Start from the Western Australia September orchid data you used in Question 1.
2.  Join it to the `tourism_total` table using `ws_id`, `year`, and `quarter`. Remember that September is in quarter 3.
3.  For Western Australia, report:
    -   the total number of September orchid sightings
    -   the average `total_trips`
    -   the proportion of sightings that have a non‑missing `max` temperature.

What do these numbers suggest about how much tourism and weather information is available for the places where orchids are being recorded?

#### Solution 2

```{r}
#| label: q2-solution
#| code-fold: true

wa_sep_q <- orchids |>
  filter(obs_state == "Western Australia",
         month == 9) |>
  mutate(quarter = 3L)

wa_sep_tour <- wa_sep_q |>
  left_join(
    tourism_total,
    by = c("ws_id", "year", "quarter")
  )

wa_sep_full <- wa_sep_tour |>
  left_join(weather, by = c("ws_id", "date"))

summary_q2 <- wa_sep_full |>
  summarise(
    n_sightings = n(),
    mean_trips = mean(total_trips, na.rm = TRUE),
    prop_with_max = mean(!is.na(max))
  )

summary_q2
```

Interactive Summary Table

```{r}
datatable(summary_q2)
```

There are 14,911 September orchid sightings in Western Australia, but the average tourism trip count at those stations is only about 33. Just under 7% of sightings have a recorded maximum temperature. This tells us that most orchid records sit in areas with relatively low reported tourism, and weather data is still very patchy for these locations.

### Question 3 — Sightings vs tourism

Now investigate whether tourism relates to recorded sightings.

Tasks:

Compute yearly statistics

Visualize the relationship

#### Solution 3

```{r}
#| label: q3-solution
#| code-fold: true

yearly_wa <- wa_sep_tour |>
  group_by(year) |>
  summarise(
    n_sightings = n(),
    mean_trips = mean(total_trips, na.rm = TRUE),
    .groups = "drop"
  )

yearly_wa
```

Interactive Scatter Plot

```{r}
plot_ly(
  yearly_wa,
  x = ~mean_trips,
  y = ~n_sightings,
  type = "scatter",
  mode = "markers+lines",
  text = ~year
) |>
  layout(
    title = "Tourism vs Orchid Sightings",
    xaxis = list(title = "Average tourism trips"),
    yaxis = list(title = "Orchid sightings")
  )
```

Overall, there appears to be a weak positive relationship between tourism and orchid sightings. Years with higher average tourism trip counts tend to show more recorded September orchid sightings in Western Australia. However, this pattern is likely explained by observer effort rather than causation: when more tourists visit an area, there are simply more people available to observe and record orchids, which increases the number of reported sightings.

\## Finishing up

To explore further, you might:

-   Re-run Question 1 using `inner_join()` instead of `left_join()` when adding tourism data. How do your counts change, and why?
-   Try a different state or month (e.g., New South Wales in August). Do you see similar gaps in weather and tourism coverage?
-   Replace the plotly scatter with a faceted `ggplot2` view by year to compare with the interactive version.
